Efficient pedigree recording for fast population genetics simulation

### 2018-06-03 review-responses.tex

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it PLoS Computational Biology}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are pleased to submit a revision of our manuscript,

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
    Jerome Kelleher, Kevin Thornton, \\
    Jaime Ashander \emph{and} Peter Ralph \\
}\\
\end{minipage}

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
    Both reviewers are clearly very enthusiastic about this manuscript, but both
    have made suggestions that require major revisions to address. Please be
    very careful in addressing the following points exhaustively:
    \begin{itemize}
        \item  Discuss the limitations imposed by Wright-Fisher assumptions (Rev 1) 
            and how the proposed approach may/may not be extended 
            to other demographic and population scenarios.

        \item  Improve the readability of the text and provide necessary background information (Rev 1)

        \item  Discuss implementation availability (Rev 2)
    \end{itemize}
\end{quote}

We thank the reviewers for helpful feedback, and respond in detail to their points below.
Something we did that should help address all these points
was to make more clear the main purpose of this paper:
to describe algorithms and data structures for tree sequences,
and how these are useful in simulation.
Our main goal is not to write a simulation package.
This is useful and important, because: 
(a) we provide a fully functional and flexible, well-documented API
so that developers can include this idea in their own simulator
(with no restrictions beyond conserved synteny of chromosomes); and
(b) we explain the data structures,
which are closely tied to the idea of a tree sequence,
and therefore important for even end users of a simulator to understand.
We describe and implement Wright--Fisher simulations not because of any limitation of our method,
but to keep things as simple as possible, only focusing on those aspects necessary 
for recording the tree sequence.

TODO: mention the tutorials?

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    Kelleher and colleagues present a novel method to efficiently record pedigree
    information in forward-time individual-based simulations of the Wright-Fisher
    process in a single large population of haploids. The method shows a large
    performance improvement compared to forward-time simulations of the same
    Wright-Fisher model without pedigree recording. The gain obtained with pedigree
    recording is due to the fact that mutations are generated post-hoc by placing
    them on the gene trees that resulted from the forward-time demographic process
    recorded as pedigree and cross-over information. The method for storing
    "succinct tree sequences" has been implemented in the coalescent-simulation
    software msprime and the authors here provide a Python API to use the data
    format previously developed in msprime in forward-time simulators. This
    represents a neat addition to the tool kit of population genetics simulations.
\end{quote}

We'd like to emphasize that although the tools for dealing with tree sequences
are currently bundled with the coalescent simulator, \msprime{},
this is for historical reasons only, 
and we are in the process of separating out these tools into a separate package,
to be called \tskit.
The format for storing tree sequences was developed for the coalescent simulator,
but the work we describe here generalized the data structure and algorithms
to allow it to be used in any simulator.
We have stated this more clearly in the text at \llname{msprime_rel}.

TODO: think about switching to \tskit{} in the paper?

\begin{point}{}
    Overall, I find the method very clever and appealing, and I am convinced that
    it will be well received by the community. However, nowadays, researchers
    generate and use genomic data in organisms other than humans and from wild
    populations that by far are not Wright-Fisher populations. How are you selling
    your method to that broader category of researchers? Compared to SPLATCHE,
    which simulates structured populations with gene flow, it may look like a
    regression (quantiNemo implemented something similar, but I don't think it was
    ever published).
    \emph{[\ldots]}
    As a developer, I am interested in the limitations and the constraints that may
    prevent me to implement your method to model populations that are not W-F.
    populations. In particular, you didn't address the question of structured
    populations with gene flow, or the effect of overlapping generations. Although
    developing an approach that is efficient in structured populations is probably
    out of scope here, giving a minimum of guidelines or just discussing the issue
    would greatly improve the scope of the manuscript.
\end{point}

\reply{
    We're sorry to have not made this more clear,
    but our goal was \emph{not} to make a faster WF simulator,
    but rather to describe a scheme that can be easily applied to \emph{any} simulator
    whatsoever, as long as it has the notion of a recombining chromosome.
    We have added a note to this effect before the proof-of-concept WF algorithm \llname{ll:not_wf};
    and substantially expanded the discussion of what steps need to be taken in general \revref.
    Also see the summarizing paragraph at \llname{ll:other_words}.
    The only thing that should stand in your way as a developer
    to implementing this is bookkeeping;
    and we've provided tools to make this bookkeeping easier
    (like the \texttt{sort tables} functionality,
    so that the simulator does not have to guarentee sortedness of the tables itself).
}

\begin{point}{}
    More generally, I found the ms very technical and addressed to a specialized
    audience. I would not assume, as you do, that readers are necessarily
    proficient in coalescent theory. Therefore, please improve the introduction by
    providing more information on: \emph{(points below)}
\end{point}

\reply{
    Thanks for the suggestion;
    since this paper is entirely about forwards simulation,
    we hope not to depend on coalescent theory at all in any crucial way.
    We have added some more explanation of the relationship to simplification
    at \revref;
    and have tried to better signpost the technical arguments at \llname{ll:coal_signpost}.
}

\begin{point}{}
    The idea of running simulations forward in time for the demography and then
    backward in time for the neutral genetics is by far not new and was already
    implemented in the software SPLATCHE (Ray, Currat, Foll, Excoffier,
    Bioinformatics 2010, 26: 2993-2994), although probably not as efficiently. The
    reference should be added.
\end{point}

\reply{
    We are assuming that you are drawing the analogy of tree sequence simplification 
    to the forward-backward scheme used by SPLATCHE.
    This is a good analogy, since simplification is tracing ancestry back through time
    as does the coalescent.
    However, we think it's fair to say that it differs fundamentally:
    the coalescent is a stochastic process that is dual to the forwards-time process,
    and relies on certain assumptions for correctness;
    whereas tree sequence simplification is deterministic (just ``bookkeeping'').
    We have added something about SPLATCHE2 to the introduction \revref.
}

\begin{point}{}
  What is the use of gene trees? What can be inferred/calculated from them that
  we cannot get from the endpoint genetic polymorphism data? (An example comes
very late in the ms, should be moved up.)
\end{point}

\reply{
    We agree this makes nice motivation,
    but think it is somewhat outside of the scope of this paper:
    for our purposes here, it suffices to explain that they help everything go faster.
    We've expanded a bit \revref.
}

\begin{point}{}
  Spell out once what is meant with $\theta = 4N_e\mu$ (and make clear that $\mu$ is
  the num loci $\times$ mutation rate)
\end{point}

\reply{
    We agree this was out of place; but rather than explain, we have removed this. \revref
}

\begin{point}{}
  Make clear how to transfer from haploid to diploid case (I guess by having
  about twice the data recorded).
\end{point}

\reply{
    We've tried to make this more clear. \revref
}

\begin{point}{}
  Figure 3 doesn't show the red stars mentioned in the main text (p8).
\end{point}

\reply{
    These are visible for us; there may have been a problem with PDF transparency;
    we've set opacity on these to 100\%; let us know if they are still invisible?
}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
    This manuscript presents algorithms that could very well change the way
    population genetic simulations are done. The current methods are either
    coalescent simulations (which build trees backwards in time, but for all
    intents and purposes do not allow selection) or forward simulations (which are
    as flexible as the user desires, but comes at the cost of being slow). The
    algorithms here are super creative solutions that allow the user to simulate a
    sequence tree forward in time. This approach can dramatically shrink the
    computational time needed to do large-scale simulations. The API proposed goes
    one step further. It is based on the msprime architecture, so a user could
    ostensibly seamlessly switch from coalescent to forward simulations very easily
    (or stitch the two together to generate ancient history using the coalescent
    and then forward simulate based on the output). This really could be a game
    changing paper for the population genetics world (as well as possibly others,
    as noted in the manuscript).

    I've thought quite a bit about how to record trees in a forward simulation, and
    could only come up with ways that drastically exploded the run time. The
    approach presented here is very creative.

    In addition, the use of the compressed format of the sequence tree is quite
    amazing. The 144k over the VCF equivalent is unbelievable.
\end{quote}

Thanks!

\begin{point}{}
    I only see one major criticism: accessibility. The authors do not provide a
    clear overview of how to obtain/install/use the algorithms presented. A
    ``proof-of-concept implementation'' is really not sufficient. In contrast,
    msprime has great documentation, but the current website does not seem to
    describe the forward simulation model (in contrast to what is suggested by the
    manuscript). \url{https://github.com/ashander/ftprime} does have installation
    instructions, but seems to be entirely geared toward integration with simuPOP.
    Most of the figures are generated with fwdpp, which are supposedly included
    here (\url{https://github.com/petrelharp/ftprime_ms}), but it is not at all clear how
    one would access the code given the current organization of this repository. It
    is unrealistic to expect all downstream users to code up the algorithms
    themselves (and would result in code that is not as well tested). The authors
    should have a more dedicated repository for users to access the software and
    start running their own simulations.
\end{point}

\reply{
    As the reviewer suggests, our main goal here is to provide easy access
    to the \emph{algorithm}, not to a forward simulation model --
    we aren't describing a forward simulation model, 
    but rather a way to speed up and get more information out of an existing one.
    But, is the algorithm accessible?
    An interface to our algorithms is available at the python level or at the C level.
    The python interface is documented through the \msprime{} documentation
    (but you need to select ``latest'': \url{https://msprime.readthedocs.io/en/latest/});
    and closely mirrors the underlying C library.
    We have also written documentation at \plrXXX LINK TO TUTORIALS}.
}

\begin{point}{}
Figure 1 shows that pedigree recording seems to scale poorly in N. Going from
N=1e3 to N=1e4 results in what seems like a $\sim 5 \times$ increase in runtime, but going
up to N=5e4 results in a $> 10 \times$ increase in runtime. Given your attention to
scaling, is this expected?
\end{point}

\reply{
    This is actually the fault of a quadratic algorithm for counting mutations
    under the hood in \fwdpp{}.
    In fact, some of us (KRT) are currently using ideas from tree sequences to fix this bottleneck.
}

%%%  Minor comments:

\begin{point}{}
The word "thanks” in the discussion of coalescent caveats is awkward.
\end{point}

\reply{
    Thanks; fixed. \revref
}

\begin{point}{}
The Appendices seem to be out of order with the text. The current order does
not seem any more logical.
\end{point}

\reply{
    Good point. Fixed.
}

\begin{point}{}
In Figure 1, it would potentially be helpful to see y-axis on a log scale so
that we could see scaling on the low end. Legend for this figure refers to
``right column of panels''. How many threads were used to obtain the dashed line?
\end{point}

\reply{
    Thanks for the suggestions.
    \plr{TODO: I've stuck in the "log" figure. What do you think?}
    \plr{Kevin: how many threads?}
}

\begin{point}{}
"its'" on page 11.
\end{point}

\reply{
    Thanks. Fixed, in two places.
}

\begin{point}{}
I appreciate the ``how does it perform in theory?'' section, but the final note
on the use of ``nedigree'' seems unnecessary. If you are going to use it (which I
do not recommend), you can justify it, otherwise drop it.
\end{point}

\reply{
    We have removed this section.
}

\begin{point}{}
What characteristics of each simulated mutation are stored? On page 10, it
suggests you store the node in which they first occur, the derived state, and
the genomic position. However, the selection coefficient and dominance
coefficient? Are there other features that can be stored (e.g., nucleotide
context)?
\end{point}

\reply{
    Good question;
    the answer is that anything can be stored (as arbitrary bytes);
    we've added a note about this. \revref
}

\begin{point}{}
The code for the Python implementation of simplify is included as supplementary
information. This is entirely unnecessary, as this code should be easily
accessible from the appropriate github repository.
\end{point}

\reply{
    Perhaps, but since github does not provide a permanent home,
    and a working implementation is, arguably, the most clear explanation of an algorithm,
    we'd prefer to keep this in as supplementary material.
}

### 2018-06-03 forwards_paper.tex
responses

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath, amssymb}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{listings}
\bibliographystyle{plainnat}
% \usepackage[T1]{fontenc}

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\justc}{\texttt{C}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\priorityq}[1]{\mbox{\textbf{PriorityQueue}}({#1})}
\newcommand{\listnew}[1]{\mbox{\textbf{List}}(#1)}
\newcommand{\pqpush}[2]{#1.\mbox{\textbf{push}}(#2)}
\newcommand{\pqpop}[1]{#1.\mbox{\textbf{pop}}()}
\newcommand{\listappend}[2]{#1.\mbox{\textbf{append}}(#2)}
\newcommand{\ancsegment}[1]{\mbox{\textbf{Segment}}(#1)}
\newcommand{\taddrow}[2]{#1.\mbox{\textbf{addrow}}(#2)}
\newcommand{\attrparent}[1]{#1.\mbox{parent}}
\newcommand{\attrchild}[1]{#1.\mbox{child}}
\newcommand{\attrleft}[1]{#1.\mbox{left}}
\newcommand{\attrright}[1]{#1.\mbox{right}}
\newcommand{\attrnode}[1]{#1.\mbox{node}}
% mbox to avoid hyphenating
\newcommand{\tsimplify}[1]{\mbox{\textbf{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\textbf{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\textbf{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\textbf{EdgeTable}}(#1)}

@@

@@

\begin{abstract}
    In this paper we describe how to
    efficiently record the entire genetic history of a population
    during a forwards-time, individual-based population genetics simulation.
    This dramatically reduces the computational burden of tracking individual genomes
    by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording history as a \emph{succinct tree sequence} as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    We implement the method in two popular forwards-time simulation frameworks,
    which speeds up large, whole-genome simulations by one to two orders of magnitude.
    In addition to speed, the method has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
    can be stored in $O(N \log N + M)$ space, making it feasible to store a simulation's entire final generation (and its' history)
    rather than only a subset.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation of deep history.
    To make these possible,
    we introduce the \msprime{} Tables API, which allows for the efficient interchange
    of tree sequences between modular program components,
    and provide an algorithm to quickly simplify a tree sequence by removing history irrelevant to a given set of genomes.
\end{abstract}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g., $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of
(a) random mating,
(b) neutrality,
and (c) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent \citep{hudson1990gene, Neuhauser1997-nn},
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical
space~\citep{barton2010new,kelleher2014coalescent}
and/or have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, density-dependent effects on fitness, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
The third assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples approaching the size of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
\fwdpp{} \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\sim 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

@@

In this paper, we describe a storage method for \emph{succinct tree sequences}
(and hence, genome sequence) as well as an algorithm for simplifying these.
The data structure is \emph{succinct} in the sense that its' space usage is close to optimal,
while still allowing efficient retrieval of information (see, e.g.,
\citet{gog2014theory}).
The storage method and simplification algorithm are both implemented
as improvements to the algorithmic tools and data structures in \msprime{}.
We also describe how these tools can efficiently record,
and later process, the embellished population pedigree from a forwards-time simulation.
While providing substantial savings in computational time and space, our methods provide much
more information than simply simulating the genomes -- the tree sequence
encodes all marginal genealogies of individuals living at the end of the simulation.
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.

@@

\begin{figure}
    \includegraphics[]{sims/rawspeed}
    \caption{
        \label{fig:runtimes_selection}
    Total run time per single simulation replicate as a function of region length.
    % measured as the scaled recombination parameter $\rho = 4Nr$.
    Line color represents different diploid population sizes ($N$).
    The left figure shows run times for standard simulations including neutral mutations.
    The right column of panels shows run times of simulations that recorded the pedigree
    and added neutral mutations afterwards.
    The dashed line in the right panel shows results for an implementation using \fwdpy{}
    where the pedigree simplification steps were handled in a separate thread of execution
    and fitness calculations were parallelized.
    Simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
    }
\end{figure}

@@

To record the genealogical history of a forwards time simulation,
we need to record two things for each new chromosome:
the birth time; and the endpoints and parental IDs of each distinctly inherited segment.
These are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
To demonstrate the idea, we write out in pseudocode how to run a neutral Wright--Fisher simulation
that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length $L$.
For simplicity, we sample exactly one crossover per generation.

@@

\paragraph{Essential steps in tree sequence recording}
To record a tree sequence, including mutations,
a simulator must record:
\begin{itemize}
    \item the birth time of each individual in the Node Table,
    \item the genetic segments each individual inherits from its parents in the Edge Table, and
    \item the individual each new mutation occurs in, and the derived state in the Mutation Table,
        adding also to the Site Table if the site is previously unmutated.
\end{itemize}
Recall that here we mean ``\emph{haploid} individual'', i.e., gamete.
Before simplification, time must be translated to ``time ago'' if it is not already.
(This was avoided in Algorithm~\algref{W} since there $t$ denoted
``time until the end of the simulation''.)
Furthermore, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
being relabeled in \msprime{}'s tables as $0, 1, \ldots, N-1$,
there must be bookkeeping that keeps in sync the individual IDs as recorded by the simulator
with the node IDs recorded in the tables.
% Finally, the first step in simplification is to \emph{sort} the tables
% according to several criteria, for algorithmic efficiency,
% as described in the documentation.

@@

\paragraph{A final note:}
in preparing this manuscript,
we debated a number of possible terms for the embellished pedigree,
i.e., the ``pedigree with ancestral recombination information''.
Etymological consensus \citep{liberman2014little} has
``pedigree'' derived from the french ``pied de grue'' for the foot of a crane
(whose branching pattern resembles the bifurcation of a single parent-offspring relationship).
An analogous term for the embellished pedigree might then be \emph{nedigree},
from ``nid de grue'',
as the nest of a crane is a large jumble of (forking) branches.
We thought it unwise to use this term throughout the manuscript,
but perhaps it will prove useful elsewhere.

@@

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath, amssymb}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{listings}
% \usepackage[T1]{fontenc}

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\justc}{\texttt{C}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\priorityq}[1]{\mbox{\textbf{PriorityQueue}}({#1})}
\newcommand{\listnew}[1]{\mbox{\textbf{List}}(#1)}
\newcommand{\pqpush}[2]{#1.\mbox{\textbf{push}}(#2)}
\newcommand{\pqpop}[1]{#1.\mbox{\textbf{pop}}()}
\newcommand{\listappend}[2]{#1.\mbox{\textbf{append}}(#2)}
\newcommand{\ancsegment}[1]{\mbox{\textbf{Segment}}(#1)}
\newcommand{\taddrow}[2]{#1.\mbox{\textbf{addrow}}(#2)}
\newcommand{\attrparent}[1]{#1.\mbox{parent}}
\newcommand{\attrchild}[1]{#1.\mbox{child}}
\newcommand{\attrleft}[1]{#1.\mbox{left}}
\newcommand{\attrright}[1]{#1.\mbox{right}}
\newcommand{\attrnode}[1]{#1.\mbox{node}}
% mbox to avoid hyphenating
\newcommand{\tsimplify}[1]{\mbox{\textbf{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\textbf{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\textbf{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\textbf{EdgeTable}}(#1)}

@@

%% for submitting with responses to reviews
\newif\ifsubmission
% \submissiontrue
\submissionfalse
\ifsubmission
    % % stuff for submission
    \usepackage[nofiglist,nomarkers,figuresonly]{endfloat}
    \usepackage{setspace}
    % \doublespacing
    \bibliographystyle{plos2015}
    \newcommand\citet{\cite}
    \newcommand\citep{\cite}
    \renewcommand\includegraphics{\relax}
    \renewcommand{\includegraphics}[2][]{\relax}
    \renewcommand{\figurename}{Fig.}
    \newcommand{\Figure}{Fig }
    \newcommand{\Figures}{Fig }
\else
    % for the nice version
    \usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
    \bibliographystyle{plainnat}
    \newcommand{\Figure}{{Figure }}
    \newcommand{\Figures}{{Figures }}
\fi
% grrr
\newcommand{\citetx}[2][]{\ifsubmission#1 \cite{#2}\else\citet{#2}\fi}

% Responses to reviews
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\newcommand{\responsefile}{review-responses}  % name of the review responses file

@@

\begin{abstract}
    In this paper we describe how to efficiently record the entire genetic history of a population
    in forwards-time, individual-based population genetics simulations with arbitrary breeding models,
    population structure and demography.
    This approach dramatically reduces the computational burden of tracking individual genomes
    by allowing us to simulate only those loci that may affect reproduction
    (those having non-neutral variants).
    The genetic history of the population is recorded as a succinct tree sequence
    as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    Recording the results of each breeding event
    results in a linear growth in the amount of information that must be stored.
    However, there is a great deal of redundancy in this information,
    since the majority of individuals simulated over the course of the simulation
    will leave no descendents in the eventual sample.
    We solve this storage problem by providing an algorithm to quickly `simplify'
    a tree sequence by removing this irrelevant history for a given set of genomes.
    By periodically simplifying the history with respect to the extant population,
    we show that the total storage space required is modest
    and overall large efficiency gains can be made over classical forward-time simulations.
    We implement a general-purpose framework for recording and simplifying genealogical data,
    which can be used to make simulations of any population model more efficient.
    We modify two popular forwards-time simulation frameworks to use this new approach
    and observe efficiency gains in large, whole-genome simulations of one to two orders of magnitude.
    In addition to speed, our method for recording pedigrees has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
        can be stored in $O(N log N + M)$ space,
        making it feasible to store a simulation's entire final generation
        (and its history) rather than only a subset.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation 
        of deep history.
\end{abstract}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees}
and thus bringing within-population trees into better focus;
second,
producing analytical methods to estimate parameters of interest from genetic data; \revpoint{1}{5}
and finally, providing a computationally feasible method
to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
under the following important assumptions: \revpoint{2}{3}
(a) random mating,
(b) neutrality,
(c) large population size,
and (d) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent \citep{hudson1990gene, Neuhauser1997-nn},
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical
space~\citep{barton2010new,kelleher2014coalescent}
and/or have large numbers of loci under various sorts of selection.
Usually, the relationship between the life history of a species --
fecundity and mortality schedules, density-dependent effects on fitness, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
More mechanistic models are possible using ``forwards--backwards'' simulations,
that first simulate population size changes forwards in time
and then thread a coalescent backwards \citep{ray2010splatche2}, \revpoint{1}{3}
but these still require the assumptions above to be met for each subpopulation.
The last assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples approaching the size of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
\fwdpp{} \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\sim 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

@@

In this paper, we describe a storage method for \emph{succinct tree sequences}
(and hence, genome sequence) as well as an algorithm for simplifying these.
The data structure is \emph{succinct} in the sense that its space usage is close to optimal,
while still allowing efficient retrieval of information (see, e.g., \citet{gog2014theory}).
We also describe how these tools can efficiently record,
and later process, the embellished population pedigree from a forwards-time simulation.
While providing substantial savings in computational time and space, our methods provide 
in principle much more information than simply simulating the genomes -- the tree sequence
encodes all marginal genealogies of individuals living at the end of the simulation.
These marginal genealogies enable fast data storage and processing,
but also provide additional information that can be used to better understand
the notoriously complex dynamics of population genetics. \revpoint{1}{4}
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.
This work originated as improvements to the algorithmic tools and data structures 
in the coalescent simulator \msprime{}; \linelabel{msprime_rel}
the tools are currently available bundled with the Python package \msprime{},
but will soon be separately available as a Python package named \tskit{}.

@@

\begin{figure}
    \includegraphics[]{sims/rawspeed_logy}
    \caption{
        \label{fig:runtimes_selection}
    Total run time per single simulation replicate as a function of region length.
    % measured as the scaled recombination parameter $\rho = 4Nr$.
    Line color represents different diploid population sizes ($N$).
    The left figure shows run times for standard simulations including neutral mutations.
    The right column shows run times of simulations that recorded the pedigree
    and added neutral mutations afterwards.
    The dashed line in the right panel shows results for an implementation using \fwdpy{}
    where the pedigree simplification steps were handled in a separate thread of execution
    and fitness calculations were parallelized.
    Simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
    }
\end{figure}

@@

To record the genealogical history of a forwards time simulation,
we need to record two things for each new chromosome:
the birth time; and the endpoints and parental IDs of each distinctly inherited segment.
These are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
To demonstrate the idea, we write out in pseudocode how to run a neutral Wright--Fisher simulation
that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length $L$.
For simplicity, we sample exactly one crossover per generation.
Note that the \emph{table recording} portion of the algorithm does not depend on the Wright--Fisher
nature of the population simulation;  \linelabel{ll:not_wf}
next we will describe how to record tables from \emph{any} simulation.

@@

\paragraph{Essential steps in tree sequence recording}
Since a tree sequence can record the history of genetic inheritance in any situation
(requiring only unambiguous inheritance and no time travel),
any individual-based population genetics simulator can maintain a tree sequence
with only a little bookkeeping. \revpoint{1}{1}
We have furthermore provided several tools to minimize this bookkeeping,
only requiring a little one-way output at the birth of each new individual.
Concretely, to record a tree sequence, including mutations,
a simulator must record for each new genome
(so, twice for each new diploid individual): \revpoint{1}{6}
\begin{itemize}
    \item the birth time of the genome in the Node Table,
    \item the segments the genome inherits from its parental genomes in the Edge Table,
    \item the locations of any new mutations in the Site Table,
    \item and the derived state of these new mutations in the Mutation Table
        (as well as the identiy of this genome the mutations appeared in).
\end{itemize}
Each of these can be simply appended to the ends of the respective tables.
Besides this, simplification should be run every once in a while (e.g., every 100 generations).
Before simplification, time in the Node Table must be translated to ``time ago'' if it is not already.
(This was avoided in Algorithm~\algref{W} since there $t$ denoted
``time until the end of the simulation''.)
There are also several ``cleaning'' steps, for which we provide functions in the Tables API:
\emph{sorting} according to several criteria for algorithmic efficiency;
and removing any duplicate sites from the Site Table.
After simplification, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
being relabeled in \msprime{}'s tables as $0, 1, \ldots, N-1$,
there must be bookkeeping that keeps in sync the individual IDs as recorded by the simulator
with the node IDs recorded in the tables.

In other words, \linelabel{ll:other_words}
to record a tree sequence, a simulation needs only to know
(a) which genomes recombined to produce each new genome, and how;
(b) the locations and results of any new mutations on each genome;
and (c) the identities of every currently alive individual at each time simplification occurs.

\paragraph{Storing metadata}
Applications may also want to store more information not fitting into an existing column of the tables,
such as the selection coefficient of a mutation, or the sex of an individual.
This (and, indeed, arbitrary information) can be stored in the \texttt{metadata} columns
present in Node, Site, and Mutation tables. \revpoint{2}{8}

@@

% \paragraph{A final note:}
% in preparing this manuscript,
% we debated a number of possible terms for the embellished pedigree,
% i.e., the ``pedigree with ancestral recombination information''.
% Etymological consensus \citep{liberman2014little} has
% ``pedigree'' derived from the french ``pied de grue'' for the foot of a crane
% (whose branching pattern resembles the bifurcation of a single parent-offspring relationship).
% An analogous term for the embellished pedigree might then be \emph{nedigree},
% from ``nid de grue'',
% as the nest of a crane is a large jumble of (forking) branches.
% We thought it unwise to use this term throughout the manuscript,
% but perhaps it will prove useful elsewhere.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Timing and memory use with no selection}
\label{ss:timing_nosel}
\renewcommand{\thefigure}{D\arabic{figure}}
\setcounter{figure}{0}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Memory usage with selection}
\label{ss:memuse}
\renewcommand{\thefigure}{B\arabic{figure}}
\setcounter{figure}{0}
